Almost Orthogonal vectors

### 2014-11-26 project3/paper/paper.tex
More changes to paper

\section{Introduction}
Given a dimension $n$ and a number $N > n$ of vectors, we want to find vectors $v_1, v_2, \ldots, v_N \in \mathcal{R}^n$ of unit length, such that
$$\epsilon = \max_{i \neq j} \{ |v_i \cdot v_j|^2 \}$$
is as small as possible.

@@

In addition, we see that for $N=3$, the quantity $\epsilon$ for three unit-length vectors $v_1, v_2, v_3 \in \mathcal{R}^2$ is given by $$\max \{ |v_1 \cdot v_2|^2, |v_1 \cdot v_3|^2, |v_2 \cdot v_3|^2 \}$$

Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathcal{R}^2$ to be the reverses of $v_1, v_2, v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because if the vector $\mathcal{R}(v)$ is the reverse of the vector $v$, then $|v \cdot v'| = |\mathcal{R}(v) \cdot v'|$ for some arbitrary vector $v'$.

@@

We split our proof into two cases,
\begin{itemize}
\item \textbf{$\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$:} This is easy to see from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.

\item One of $\theta_1, \theta_2, \dots, \theta_n$ is greater than $\pi/2$:

Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$. Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$ (since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$).

@@

Figure $1$ shows the optimal configuration of vectors $v_1$, $v_2$ and $v_3$.
\begin{figure}
    \centering
    \begin{tikzpicture}[yscale=-1] 
        % x-axis
        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
        % y-axis
        \draw [thick,->] (0, 4.5) -- (0, -4.5);
        % origin label
        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
        % x-axis label
        \node at (4.5, 0.5) {\text{$x$}};
        % y-axis label
        \node at (0, -5) {\text{$y$}};
        % circle
        \draw (0,0) circle (3cm);
        \draw (3,0)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, 2.59)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);

>>>>

\section{Introduction}
Given a dimension $n$ and a number $N > n$ of vectors, we want to find vectors $v_1, v_2, \ldots, v_N \in \mathbb{R}^n$ of unit length, such that
$$\epsilon = \max_{i \neq j} \{ |v_i \cdot v_j|^2 \}$$
is as small as possible.

@@

In addition, we see that for $N=3$, the quantity $\epsilon$ for three unit-length vectors $v_1, v_2, v_3 \in \mathbb{R}^2$ is given by $$\max \{ |v_1 \cdot v_2|^2, |v_1 \cdot v_3|^2, |v_2 \cdot v_3|^2 \}$$

Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathbb{R}^2$ to be the reverses of $v_1, v_2, v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because if the vector $\mathcal{R}(v)$ is the reverse of the vector $v$, then $|v \cdot v'| = |\mathcal{R}(v) \cdot v'|$ for some arbitrary vector $v'$.

@@

We split our proof into two cases,
\begin{itemize}
\item $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$: This is easy to see from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.

\item One of $\theta_1, \theta_2, \dots, \theta_n$ is greater than $\pi/2$:
\begin{figure}
    \centering
    \begin{tikzpicture}[yscale=-1] 
        % x-axis
        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
        % y-axis
        \draw [thick,->] (0, 4.5) -- (0, -4.5);
        % origin label
        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
        % x-axis label
        \node at (4.5, 0.5) {\text{$x$}};
        % y-axis label
        \node at (0, -5) {\text{$y$}};
        % circle
        \draw (0,0) circle (3cm);
        \draw (3,0)[blue,fill=blue] circle (0.1cm);
        \draw (2.42, -1.76)[blue,fill=blue] circle (0.1cm);
        \draw (0.93, -2.85)[blue,fill=blue] circle (0.1cm);
        \draw (-2.42, -1.76)[blue,fill=blue] circle (0.1cm);
        \draw (-0.93, -2.85)[blue,fill=blue] circle (0.1cm);

        \draw [thick,-,blue] (3, 0) -- (0, 0);
        \draw [thick,-,blue] (2.42, -1.76) -- (0, 0);
        \draw [thick,-,blue] (0.93, -2.85) -- (0, 0);
        \draw [thick,-,blue] (-2.42, -1.76) -- (0, 0);
        \draw [thick,-,blue] (-0.93, -2.85) -- (0, 0);

        \node at (4, -0.3) {\text{$x_1 = [1,0]$}};
        \node at (4.5, -1.76) {\text{$x_2 = [\cos \pi/5,\sin \pi/5]$}};
        \node at (3.43, -3) {\text{$x_3 = [\cos 2\pi/5,\sin 2\pi/5]$}};
         \node at (-4.5, -1.76) {\text{$x_4 = [\cos 4\pi/5,\sin 4\pi/5]$}};
        \node at (-3.43, -3) {\text{$x_5 = [\cos 3\pi/5,\sin 3\pi/5]$}};
    \end{tikzpicture}
    \caption{A configuration of five unit vectors that produce the optimum $\epsilon$ for $n=2, N=5$}
\end{figure}

Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$. Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$. (since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$)

It is easy to see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\epsilon$ is equal to $\cos^2 \theta_n$.

@@

From the above lemma, we can conclude that if $j = \text{argmin } \theta_i$ and $\theta_1 + \theta_2 + \theta_3 = \pi$, then $\epsilon = \cos^2 \theta_j$.

Furthermore, we can obtain an upper bound on $\theta_j$ by observing that $\theta_ 1 + \theta_ 2 + \theta_3 \geq \theta_ j + \theta_j + \theta_j = 3 \theta_j \Rightarrow \theta_j \leq \pi/3$. Since we're interested in the smallest such $\epsilon$ and since the cosine function is a decreasing function in $\theta$ between $0$ and $\pi/2$, we conclude that the optimum value of $\epsilon$ for $n=2$ and $N=3$ is $\cos^2 \pi / 3$. Figure $1$ shows this optimum configuration.

\begin{theorem}
For $n=2$ and arbitrary $N$, the optimum value of $\epsilon$ is given by $\cos^2 \pi/N$.
\end{theorem}

\begin{proof}
The proof is easy to see given the above lemma.

Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$.

Let us define $\theta_i$ as the angle between the vectors $x_i$ and $x_{i+1}$. Then, we quickly see that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$. If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$.

Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$.
\end{proof}

Figure $2$ shows an optimum configuration for $n=2$ and $N=5$.
\begin{figure}
    \centering
    \begin{tikzpicture}[yscale=-1] 
        % x-axis
        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
        % y-axis
        \draw [thick,->] (0, 4.5) -- (0, -4.5);
        % origin label
        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
        % x-axis label
        \node at (4.5, 0.5) {\text{$x$}};
        % y-axis label
        \node at (0, -5) {\text{$y$}};
        % circle
        \draw (0,0) circle (3cm);
        \draw (3,0)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, 2.59)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);
